A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-13 main.tex
Comments on term equivalence.

As usual in dependent type systems, type equivalence, used in \TConv,
is important.  The type equivalence judgment of the form
$\G \V \tau \E \sigma : K @ A$ means that types $\tau$ and $\sigma$
are equivalent as types of kind $K$ at stage $A$ under $\G$.
Figure~\ref{fig:type-equivalence-rules} shows the rules for type
equivalence.  Type equivalence is basically the least congruence
closed under term equivalence.  The ules for compatibility (closure
under type formation) are derived from corresponding typing rules in a
straightforward manner.  The rules are a little simpler than some
dependent type systems \AI{such as?}  because there is no abstraction
at the type level.

@@

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A}{\QTAbs } \\[2mm]
		\infrule{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A}{\QTApp } \\[2mm]
		\infrule{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}{\textsc{QT-$\TW$}}\andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}}{\textsc{QT-Csp}} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A}{\textsc{QT-Gen}} \\[2mm]
		\infrule{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A}{\textsc{QT-Refl}} \andalso
		\infrule{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A}{\textsc{QT-Sym}} \\[2mm]
		\infrule{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}{\textsc{QT-Trans}}
		\caption{Type Equivalence Rules.}
		\label{fig:type-equivalence-rules}
	\end{center}
\end{figure}

The term equivalence judgment of the form $\G \V M \E N : \rho @ A$
means that terms $M$ and $N$ are equivalent as terms of type $\rho$ at
stage $A$ under $\G$.

%%%%%%%%%%%%%%%%%%%%%%% RESUME FROM HERE %%%%%%%%%%%%%%%%%%%%%%%%%%

The equivalence of terms is defined in Figure \ref{fig:term-equivalence-rules}.
\QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns, \QCsp\ are generated directly from the syntax of a term.
\QRefl, \QSym, \QTrans\ exist in order to make the term equivalence relation an equivalence relationship as with the type equivalence.
The rule \QBeta\ means when a term $M \longrightarrow_\beta M'$, $M$ is equivalent to $M'$.
\QTBLTB and \QLambda\ are corresponding to $\longrightarrow_\blacklozenge$ and $\longrightarrow_\Lambda$, respectively.

\begin{figure}
	\begin{center}
		\infrule{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A}{\QAbs} \\[2mm]
		\infrule{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A}{\QApp} \\[2mm]
		\infrule{\G\V M \E N : \tau @{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau @A}{\QTB} \andalso
		\infrule{\G\V M \E N : \TW_\alpha \tau @A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau @{A\alpha}}{\QTBL} \\[2mm]
		\infrule{\G\V M\E N : \tau @A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau @A}{\QGen} \\[2mm]
		\infrule{\G\V M \E N:\forall\alpha.\tau @A}{\G\V M\ \epsilon \E N\ \epsilon : \tau[\alpha \mapsto \epsilon]@A}{\QIns}\andalso
		\infrule{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau @{A\alpha}}{\QCsp} \\[2mm]
		\infrule{\G\V M:\tau @A}{\G\V M\E M : \tau @A}{\QRefl} \andalso
		\infrule{\G\V M\E N : \tau @A}{\G\V N\E M : \tau @A}{\QSym} \\[2mm]
		\infrule{\G\V M\E N : \tau @A \andalso \G\V N\E L : \tau @A}{\G\V M\E L : \tau @A}{\QTrans} \\[2mm]
		\infrule{\G,x:\sigma@A\V M:\tau @A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A}{\QBeta} \\[2mm]
		% \infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
		\infrule{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau @A}{\G\V (\Lambda\alpha.M)\ \epsilon \E M[\alpha \mapsto \epsilon] : \tau[\alpha \mapsto \epsilon]@A}{\QLambda} \\[2mm]
		\infrule{\G\V M \E N : \tau @A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A}{\QTBLTB} \andalso
		\infrule{\G\V M:\tau @{A\alpha} \andalso \G\V M:\tau @A}{\G\V\%_\alpha M \E M : \tau @{A\alpha}}{\QPercent}
		\caption{Term Equivalence Rules}
		\label{fig:term-equivalence-rules}
	\end{center}
\end{figure}

% \QPercentの説明
The rule \QPercent is something special because it has no background of syntax or reductions.
But we add this rule to \LMD\ because it is essential when you use dependent types in code type.
For example, we can utilize a dependently typed multi-stage calculus to generate code for matrix multiplication.
When we know sizes of matrices before the multiplication,
we can generate efficient code for multiplication by unrolling loops.

$\text{mulmat}$ function in the following code fragment generate code for matrix multiplication.
$\text{mulmat}$ takes two integers which are the size of the multiplicand matrix and generate a code.
The last integer to decide the size of the multiplier matrix is given at runtime.
You can generate code by applying two integers to $\text{mulmat}$.
We applied 3 and 5 in the second line and got $\TW_\alpha \Pi z:\I.(\M\ z\ \%_\alpha 5) \to (\M\ \%_\alpha 5\ \%_\alpha 3) \to (\M\ z\ \%_\alpha 3)$.
But this type is difficult to combine with other code because there are two $\%_\alpha$ for CSP.
\QPercent\ states that we can erase this $\%_\alpha$ under a condition which is explained in the next paragraph.

@@

>>>>

As usual in dependent type systems, type equivalence, used in \TConv,
is important.  The type equivalence judgment of the form
$\G \V \tau \E \sigma : K @ A$ means that types $\tau$ and $\sigma$
are equivalent as types of kind $K$ at stage $A$ under $\G$.
Figure~\ref{fig:type-equivalence-rules} shows the rules for type
equivalence.  Type equivalence is basically the least congruence
closed under term equivalence.  The rules for compatibility (closure
under type formation) are derived from corresponding typing rules in a
straightforward manner.  The rules are a little simpler than some
dependent type systems \AI{such as?}  because there is no abstraction
at the type level.

@@

\begin{figure}
	\begin{center}
		\infrule[{\QTAbs }]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V \rho \E \pi :: *@A}{\G\V\Pi x:\tau.\rho \E \Pi x:\sigma.\pi :: *@A} \\[2mm]
		\infrule[\QTApp]{\G\V \tau \E \sigma :: (\Pi x:\rho.K)@A \andalso \G\V M \E N : \rho @A}{\G\V \tau\ M \E \sigma\ N :: K[x \mapsto M]@A} \\[2mm]
		\infrule[\textsc{QT-$\TW$}]{\G\V \tau \E \sigma :: *@{A\alpha}}{\G\V \TW_{\alpha} \tau \E \TW_{\alpha} \sigma :: *@A}\andalso
		\infrule[\QTCsp]{\G\V \tau \E \sigma :: K@A}{\G\V \tau \E \sigma :: K@{A\alpha}} \\[2mm]
		\infrule[\QTGen]{\G\V \tau \E \sigma :: *@A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V \forall\alpha.\tau \E  \forall\alpha.\sigma :: *@A} \\[2mm]
		\infrule[\QTRefl]{\G\V \tau::K@A}{\G\V \tau\E\tau :: K@A} \andalso
		\infrule[\QTSym]{\G\V \tau \E \sigma :: K@A}{\G\V \sigma \E \tau :: K@A} \\[2mm]
		\infrule[\QTTrans]{\G\V \tau \E \sigma :: K@A \andalso \G\V \sigma \E \rho  :: K@A}{\G\V \tau \E \rho  :: K@A}
		\caption{Type Equivalence Rules.}
		\label{fig:type-equivalence-rules}
	\end{center}
\end{figure}

The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
which means that terms $M$ and $N$ are equivalent as terms of type
$\rho$ at stage $A$ under $\G$, is defined by the rules in
Figure~\ref{fig:term-equivalence-rules}.  Most rules are
straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
\QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence; the
rules \QBeta, \QTBLTB, and \QLambda{} correspond to the reduction rules.
\AI{Don't we give names to those reduction rules?}

\begin{figure}
	\begin{center}
		\infrule[\QAbs]{\G\V \tau \E \sigma :: *@A \andalso \G,x:\tau @A \V M \E N : \rho @A}{\G\V\lambda x:\tau.M \E \lambda x:\sigma.N : (\Pi x:\tau.\rho)@A} \\[2mm]
		\infrule[\QAbs]{\G\V M \E L : (\Pi x:\sigma.\tau)@A \andalso \G\V N \E O : \sigma@A}{\G\V M\ N \E L\ O : \tau[x \mapsto N]@A} \\[2mm]
		\infrule[\QTB]{\G\V M \E N : \tau @{A\alpha}}{\G\V \TB_\alpha M \E \TB_\alpha N : \TW_\alpha \tau @A} \andalso
		\infrule[\QTBL]{\G\V M \E N : \TW_\alpha \tau @A}{\G\V \TBL_\alpha M \E \TBL_\alpha N : \tau @{A\alpha}} \\[2mm]
		\infrule[\QGen]{\G\V M\E N : \tau @A \andalso \alpha \notin \FTV(\G)\cup\FTV(A)}{\G\V \Lambda\alpha.M \E \Lambda\alpha.N : \forall\alpha.\tau @A} \\[2mm]
		\infrule[\QIns]{\G\V M \E N:\forall\alpha.\tau @A}{\G\V M\ \epsilon \E N\ \epsilon : \tau[\alpha \mapsto \epsilon]@A} \andalso
		\infrule[\QCsp]{\G\V M \E N : \tau @A}{\G\V\%_\alpha M \E \%_\alpha N : \tau @{A\alpha}} \\[2mm]
		\infrule[\QRefl]{\G\V M:\tau @A}{\G\V M\E M : \tau @A} \andalso
		\infrule[\QSym]{\G\V M\E N : \tau @A}{\G\V N\E M : \tau @A} \\[2mm]
		\infrule[\QTrans]{\G\V M\E N : \tau @A \andalso \G\V N\E L : \tau @A}{\G\V M\E L : \tau @A} \\[2mm]
		\infrule[\QBeta]{\G,x:\sigma@A\V M:\tau @A \andalso \G\V N:\sigma@A}{\G\V(\lambda x:\sigma.M)\ N\E M[x\mapsto N] : \tau[x \mapsto N]@A} \\[2mm]
		% \infrule{\G\V M:(\Pi x:\sigma.\tau)@A \andalso x\notin \text{FV}(M)}{\G\V(\lambda x:\sigma.M\ x)\E M: (\Pi x:\sigma.\tau)@A}{\QEta} \\[2mm]
		\infrule[\QLambda]{\G\V (\Lambda\alpha.M) : \forall\alpha.\tau @A}{\G\V (\Lambda\alpha.M)\ \epsilon \E M[\alpha \mapsto \epsilon] : \tau[\alpha \mapsto \epsilon]@A} \\[2mm]
		\infrule[\QTBLTB]{\G\V M \E N : \tau @A}{\G\V \TBL_\alpha(\TB_\alpha M) \E N : \tau @A} \hfil
		\infrule[\QPercent]{\G\V M:\tau @{A\alpha} \andalso \G\V M:\tau @A}{\G\V\%_\alpha M \E M : \tau @{A\alpha}}
		\caption{Term Equivalence Rules.}
		\label{fig:term-equivalence-rules}
	\end{center}
\end{figure}

% \QPercentの説明
The only rule that deserves elaboration is the last rule \QPercent.
Intuitively, it means that the CSP operator applied to term $M$ can be
removed if $M$ is also well typed at the next stage \(A\alpha\).
For example, constants do not depend on the stage (see \TConst) and
so \(\G\V \%_\alpha c \E c : \tau @ A\alpha\) holds but variables
do depend on stages and so this rule does not apply.

Interestingly, Hanada and Igarashi~\cite{} rejected the idea of
reduction to remove $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator
in implementation.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense and as we argue \AI{where?}
it is practically significant.

%%%%%%%%%%%%%%%%%%% SKIP FROM HERE %%%%%%%%%%%%%%%%%%%%
$\text{mulmat}$ function in the following code fragment generate code for matrix multiplication.
$\text{mulmat}$ takes two integers which are the size of the multiplicand matrix and generate a code.
The last integer to decide the size of the multiplier matrix is given at runtime.
You can generate code by applying two integers to $\text{mulmat}$.
We applied 3 and 5 in the second line and got $\TW_\alpha \Pi z:\I.(\M\ z\ \%_\alpha 5) \to (\M\ \%_\alpha 5\ \%_\alpha 3) \to (\M\ z\ \%_\alpha 3)$.
But this type is difficult to combine with other code because there are two $\%_\alpha$ for CSP.
\QPercent\ states that we can erase this $\%_\alpha$ under a condition which is explained in the next paragraph.

@@

%%%%%%%%%%%%%%%%%%% TO HERE %%%%%%%%%%%%%%%%%%%%
